Icicle: write once, run once

### 2016-05-27 2016fhpc/section/02-Source.tex
Change evaluator to take modality; update some benchmarks; hide conclusion

We define simple values $V$ as numbers, booleans, maps of values, or tuples of values.
Next, stream values $V'$ correspond to modalities: pure computations are represented as a single value;
@Element@s as a stream transformer from a heap to a value;
and @Aggregate@s as folds defining an initial state and update and extract functions.

@@

The judgment $\SourceStepX{E'}{V'}$ defines a big-step evaluation relation.

(EVal) applies when the expression is alread a stream value.
This corresponds, roughly, to the (TcNat) typechecking rule.

(EBoxStream) and (EBoxFold) convert a pure value to a stream and fold, respectively.
The stream transformer ignores the input stream and returns the pure value.
The fold has a unit accumulator, and the extract function simply returns the pure value.

Next (EPrimValue), (EPrimStream) and (EPrimFold) evaluate primitive applications.
For pure values, the values are unboxed and the primitive is applied.
For streams, a new stream is created and the input heap applied to all streams before applying the primitive.
Folds are the most complicated: the fold accumulators are tupled together, for example $(\times_i~z_i)$ being the tuple of all the initial accumulators.
At each step, the accumulators for all the input folds are updated with their own update function, and the input heap.
Finally, the extract function extracts the final values of the input folds, and applies the primitive.

The (ELet) rule is fairly standard, evaluating the bound to a value, substituting it in, and evaluating the rest.

(EFold) is one of the most complex rules, as it introduces recursion to the streams.
It evaluates its initial to a pure value, $z'$, which is used as the initial state for the fold.
In the update expression, the fold binding $x$ is substituted with a stream that looks up $x$ in its input heap.
This action-at-a-distance is necessary because the value of $x$ depends on the value of the fold, which is not known yet.
Finally, the return fold is constructed, which binds $x$ to the current fold accumulator.
The extract function for this is the identity, simply returning the accumulator.

(EFilter) evaluates its predicate to a stream of booleans, and the rest to a fold.
The result is a fold which only applies the update function when the predicate stream is true.

(EGroup) is rather similar to filter, except that it returns a map.
Here, we start with an empty map of bottoms.
At every update step, we find the key and look up the current accumulator for that key in the map, or the fold initialiser if it is bottom.
We then apply the update function to the accumulator, and update that key in the map.
For the extract function, we simply apply the extract to each element of the map.

In order to execute these evaluation rules on an actual input table, a few steps are necessary.
First, variables referring to column names such as @open@ or @close@ must have stream values substituted in, so that the actual value will be looked up in the input heap.
For example, $e[@open@~:=~@Stream@~(\lam{\sigma} \sigma~@open@)]$.
Next, the query must be converted to a @Fold@.
Then, for each record, the update function would be called with an input heap binding the record values: $\{@open@ = 1, @close@ = 2\}$ and so on.
This occurs for every record in the table.
Finally, the extract function is called, which gives the result of the query.

>>>>

Definition $V$ are simple values of numbers, booleans, maps and tuples.
Definition $V'$ are stream values that correspond to modalities: pure computations are represented as a single value;
@Element@s as a stream transformer from a heap to a value;
and @Aggregate@s as folds defining an initial state and update and extract functions.
Definition $M'$ are the modalities themselves without value types.

@@

The judgment $\SourceStepX{M'}{E'}{V'}$ defines a big-step evaluation relation.
The rules below use the modality in deciding whether pure values should be converted to streams or folds.

Rule EVal applies when the expression is alread a stream value.
This corresponds roughly to the TcNat typechecking rule.

Rules EBoxStream and EBoxFold convert a pure value to a stream and fold respectively, when the environment modality is not pure.
The stream transformer ignores the input stream and returns the pure value.
The fold has a unit accumulator, and the extract function simply returns the pure value.

Rules EPrimValue, EPrimStream and EPrimFold evaluate primitive applications.
For pure values, the values are unboxed and the primitive is applied.
For streams, a new stream is created and the input heap applied to all streams before applying the primitive.
Folds are the most complicated: the fold accumulators are tupled together, for example $(\times_i~z_i)$ being the tuple of all the initial accumulators.
At each step, the accumulators for all the input folds are updated with their own update function, and the input heap.
Finally, the extract function extracts the final values of the input folds, and applies the primitive.

Rule ELet rule is fairly standard, evaluating the bound to a value, substituting it in, and evaluating the rest.
The only difference is we inspect the type of the bound expression to find the modality, then evaluate the bound under that modality.
The rest of the expression is evaluated under the original modality.

Rule EFold is one of the most complex, as it introduces recursion to the streams.
It evaluates its initial to a pure value, $z'$, which is used as the initial state for the fold.
In the update expression, the fold binding $x$ is substituted with a stream that looks up $x$ in its input heap.
This sleight of hand is necessary because the value of $x$ depends on the value of the fold, which is not known yet.
Finally, the return fold is constructed, which binds $x$ to the current fold accumulator.
The extract function for this is the identity, simply returning the accumulator.

Rule EFilter evaluates its predicate to a stream of booleans, and the rest to a fold.
The result is a fold which only applies the update function when the predicate stream is true.

Rule EGroup is rather similar to filter, except that it returns a map.
Here, we start with an empty map of bottoms.
At every update step, we find the key and look up the current accumulator for that key in the map, or the fold initialiser if it is bottom.
We then apply the update function to the accumulator, and update that key in the map.
For the extract function, we simply apply the extract to each element of the map.

In order to execute these evaluation rules on an actual input table, a few steps are necessary.
First, variables referring to column names such as @open@ or @close@ must have stream values substituted in, so that the actual value will be looked up in the input heap.
For example, $e[@open@~:=~@Stream@~(\lam{\store} \store~@open@)]$.
Next, the query must be converted to a @Fold@.
Then, for each record, the update function would be called with an input heap binding the record values: $\{@open@ = 1, @close@ = 2\}$ and so on.
This occurs for every record in the table.
Finally, the extract function is called, which gives the result of the query.

### 2016-05-27 2016fhpc/section/04-Benchmarks.tex
Change evaluator to take modality; update some benchmarks; hide conclusion

We are currently using Icicle in production, over medium-sized datasets that fit on a single disk.
Initial results have been very promising, and we are currently working towards a distributed implementation for larger datasets.

One evaluation we performed was replacing a hand-written R script which had been running in production.
This R script works over around three hundred gigabytes of PSV data.
It computes twelve queries over each of the thirty-one input tables, computing 372 queries in total.

The R script for this takes around fifteen hours to run and is 3,566 lines, 2,311 of which are code.
In contrast, our Icicle queries take eleven minutes to run, and the dictionary describing the queries is 191 lines of code.

@@

\begin{code}
          R       Icicle
Minutes   900         11
LOC       2,311      191
MB/s      6.6        543
\end{code}

Run some benchmarks.

\begin{code}
        grep "^$"   wc    wc -l   Icicle
MB/s      74        251   775       1,060
\end{code}

>>>>

We are currently using Icicle in production, over medium-sized datasets (around three hundred gigabytes) that fit on a single disk.
While future versions will require incremental computation, the runtime processing medium-sized datasets has been small enough to run over the entire dataset each time.
These initial results have been very promising, and we are currently implementing distribution across multiple machines for datasets that do not fit on a single disk.
Incremental computation becomes more important for this distributed case, as it can potentially avoid copying terabytes of data over the network.

We evaluated Icicle by replacing an R script which ran weekly in production.
This R script works over around three hundred gigabytes of PSV data.
It computes twelve queries over each of the thirty-one input tables, computing 372 queries in total.

The R script for this takes fifteen hours to run and is 3,566 lines, 2,311 of which are code.
In contrast, our Icicle queries take seven minutes to run, and the dictionary describing the queries is 191 lines of code.

@@

\begin{figure}

\begin{tikzpicture}
\begin{axis}[
	x tick label style={/pgf/number format/1000 sep=},
	ylabel=MB/s,
	enlargelimits=0.01,
	legend style={at={(0.5,-0.1)},anchor=north,legend columns=-1},
	ybar interval=0.7,
  symbolic x coords={R, wc, grep, wc -l, cat, Run 372, Run 1, end}
]
\addplot coordinates {(R,6.0) (wc,182.6) (grep,261) (wc -l,852) (cat,878) (Run 372,908) (Run 1,1029) (end,0) };

% \legend{MB/s}

\end{axis}
\end{tikzpicture}

\caption{Throughput comparisons of Icicle against existing R code and standard Unix utilities}
\label{fig:bench:other}
\end{figure}

\begin{figure}

\begin{tikzpicture}
\begin{axis}[
	x tick label style={/pgf/number format/1000 sep=},
	ylabel=MB/s,
  ymin=800, ymax=1100,
%  xmax=372,
  xlabel=Number of queries,
	enlargelimits=0.01,
	legend style={legend columns=-1},
%	legend style={at={(0.5,-0.1)},anchor=north,legend columns=-1},
%	ybar interval=0.7,
]

% bench/raw.txt
% Running with fewer queries
 \addplot coordinates {(0,1025.5) (1,1029.6) (48,978) (96,941) (144,943) (192,936) (240,904) (276,895.5) (324,881) (372,846.2) };

% bench/raw.txt
% Running with fewer queries, better distribution
\addplot coordinates { (0,1029) (31,1013) (62,1006) (93,998) (124,991) (155,941) (186,956) (217,933) (248,931) (279,915) (279,915) (310,919) (341,916) (372,908) };

% Fusion, no output
\addplot coordinates { (0,1001) (31,1038) (62,1025) (93,1031) (124,1006) (155,1015) (186,1020) (217,997) (248,1026) (279,1024) (310,1009) (341,1038) (372,1025) };

\legend{CSE, No CSE, dev/null};

\end{axis}
\end{tikzpicture}

\caption{Decrease in throughput as queries are added, with and without common subexpression elimination (CSE) over queries}
\label{fig:bench:queries}
\end{figure}

The table in figure~\ref{fig:bench:other} shows the throughput in megabytes per second.
We compared the throughput of several programs over the same 317GB dataset:
\begin{itemize}
\item our original R implementation (R);
\item Icicle with equivalent queries (Run 372);
\item Icicle with a single query computing the mean (Run 1);
\item finding empty lines with @grep "^$"@;
\item counting characters, words and lines with @wc@;
\item counting only lines with @wc -l@; and
\item reading and throwing away the results with @cat > /dev/null@.
\end{itemize}

We ran all of the unix utilities with unicode disabled using @LANG=C LC_COLLATE@ for maximum performance.
The computer we used for most of these was an Amazon EC2 @c3.8xlarge@ with 32 CPUs, 60GB of RAM, and striped, RAIDed SSD storage.
The R code requires an @i2.8xlarge@ with 244GB of memory, but we were unable to perform our other benchmarks on such a large machine.
Icicle significantly outperformed R, @grep@ and @wc@, and was roughly on par with @wc -l@ and @cat@.
Although the computer had 32 CPUs, Icicle was the only benchmark we ran that used multiple CPUs: the other benchmarks were single threaded.

Figure~\ref{fig:bench:queries} shows how Icicle scales in throughput as more queries are added.
